---
title: "Multi-scale model assessment"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Multi-scale model assessment}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = rlang::is_installed("ggplot2") && rlang::is_installed("tigris")
)
```

```{r setup, include=FALSE}
ggplot2::theme_set(ggplot2::theme_minimal())
```

This vignette walks through how to use waywiser to assess model predictions at
multiple spatial scales, using the `ny_trees` data in waywiser, adapted from 
that post. 

First things first, we'll set up our environment, loading a few packages 
and telling sf to download the coordinate reference system for our data,
if needed:

```{r}
library(sf)
library(tidyr)
library(dplyr)
library(waywiser)

invisible(sf_proj_network(TRUE))
```

The data we're working with is extremely simple, reflecting the number of trees
and amount of aboveground biomass ("AGB", the total amount of aboveground woody 
bits) at a number of plots across New York State. We can plot it to see that 
there's some obvious spatial dependence in this data -- certain regions have
clusters of much higher AGB values, while other areas (such as the area around
New York City to the south) have clusters of much lower AGB.

```{r}
library(ggplot2)

ny_trees %>%
  ggplot() +
  geom_sf(aes(color = agb), alpha = 0.4) +
  scale_color_distiller(palette = "Greens", direction = 1)
```

Because our focus here is on model _assessment_, not model fitting, we're going
to use an extremely simple linear regression to try and model AGB across the 
state. We'll predict AGB as being a linear function of the number of trees at
each plot, and then we're going to use this model to predict expected AGB:

```{r}
agb_lm <- lm(agb ~ n_trees, ny_trees)
ny_trees$predicted <- predict(agb_lm, ny_trees)
```

Now we're ready to perform our multi-scale assessments. The `ww_multi_scale()`
function supports two different methods for performing assessments: first, 
you can pass arguments to `sf::st_make_grid()` (via `...`), specifying the sort
of grids that you want to make. For instance, if we wanted to make grids with
apothems (the distance from the middle of a grid cell to the middle of its 
sides) ranging from 10km to 100km long, we can call the function like this:

```{r}
cell_sizes <- seq(10, 100, 10) * 1000
ny_multi_scale <- ww_multi_scale(
  ny_trees,
  agb,
  predicted,
  cellsize = cell_sizes
)

ny_multi_scale
```

We've now got a tibble with estimates for our model's RMSE and MAE at each scale 
of aggregation! We can use this information to better understand how our model 
does when predictions are being aggregated across larger units than a single
plot; for instance, our model _generally_ does better at larger scales of 
aggregation:

```{r}
ny_multi_scale %>%
  unnest(.grid_args) %>%
  ggplot(aes(x = cellsize, y = .estimate, color = .metric)) +
  geom_line()
```

Note that we used the `.grid_args` column, which stores the arguments we used to
make the grid, to associate our performance estimates with their corresponding
`cellsize`.

In addition to our top-level performance estimates, our `ny_multi_scale` object
also includes our true and estimated AGB, aggregated to each scale, in the 
`.grid` column. This lets us easily check what our predictions look like at each 
level of aggregation:

```{r}
ny_multi_scale$.grid[[9]] %>%
  filter(!is.na(.estimate)) %>%
  ggplot(aes(fill = .estimate)) +
  geom_sf() +
  scale_fill_distiller(palette = "Greens", direction = 1)
```

In addition to specifying systematic grids via `sf::st_make_grid()`, 
`ww_multi_scale()` also allows you to provide your own aggregation units. For 
instance, we can use the `tigris` package to download census block group 
boundaries, as well as county and county subdivision boundaries, for the state 
of New York:

```{r message=FALSE, results='hide'}
suppressPackageStartupMessages(library(tigris))

ny_block_groups <- block_groups("NY")
ny_county_subdivisions <- county_subdivisions("NY")
ny_counties <- counties("NY")
```

We can then provide those `sf` objects straight to `ww_multi_scale`:

```{r warning=FALSE}
ny_division_assessment <- ww_multi_scale(
  ny_trees,
  agb,
  predicted,
  grids = list(
    ny_block_groups,
    ny_county_subdivisions,
    ny_counties
  )
)

ny_division_assessment %>%
  mutate(
    division = rep(c("Block group", "County subdivision", "County"), each = 2)
  ) %>%
  ggplot(aes(x = division, y = .estimate, fill = .metric)) +
  geom_col(position = position_dodge())
```

By providing grids directly to `ww_multi_scale()`, we can see how well our model 
performs when we aggregate predictions to more semantically meaningful levels 
than the systematic grids.
